Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2020-06-03 doc/network.tex
clean u[

At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
While this definition is widely used in the literature, it is not connected directly to a population-level \emph{distribution}.
There are important distinctions to be made when defining generation-interval distributions at the population-level. The \emph{intrinsic} generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case \citep{champredon2015intrinsic}.
On the other hand, \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
Since some infectious contacts will be made with non-susceptible people, and thus not result in infection, realized distributions can differ systematically from the intrinsic distribution.

The shape of the realized generation-interval distribution depends on the reference time and perspective \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual. 
Thus, ``backward'' generation intervals (which look at a cohort of infectees, and ask when their infectors were infected) will be shorter on average than intrinsic generation intervals -- the converse is true when an epidemic is subsiding \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
Likewise, we can define ``forward'' generation intervals, which look at a cohort of infectors and ask when their infectees were infected.
Mean forward generation intervals tend to decrease over the course of an epidemic due to susceptible depletion \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic}.

Realized generation intervals are also affected by spatial structure.
In a population that does not mix homogeneously, susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population.
This means that infectious contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already).
As a result, realized generation intervals (from the perspective of an infector) will typically have a shorter mean than the intrinsic generation-interval distribution in a non-homogeneous population.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.

In practice, realized generation intervals are often difficult to measure for many diseases, because it is difficult to observe when individuals become infected;
in most cases, observable events are clinical (e.g. onset of symptoms).
There are some exceptions: for example, generation intervals are commonly measured directly through contact tracing for canine rabies, where infection events are bites \citep{hampson2009transmission}.
Intervals between observed disease progression events (commonly, onset of signs or symptoms) are called serial intervals \citep{svensson2007note}.
Serial intervals are in many ways similar to generation intervals, but there are also complexities in their use  \citep{britton2019estimation}.
We will not address these complexities here.

While an epidemic is ongoing, realized generation intervals, at least in theory, can be measured by identifying who infected whom and when, and aggregated to form a single distribution.
We typically want to try to make inference based on this aggregated distribution -- that is, on all available data that have been gathered since the beginning of an epidemic.
These aggregated measurements are ``truncated'' because we do not know what happens after the time of last observation.
The distribution of these truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded in time to be observed.
We therefore predict that removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which contains information about the population structure and allows us to correctly infer the initial reproductive number from the initial exponential growth rate.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{individual_and_population.pdf}
\caption{\textbf{Comparison of individual- and population-level kernels.}
(Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days;
this represents an individual realization of a random process.
(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 and 5 days, respectively;
this represents a population average of a random process.
Shaded areas under the curves are equal to individual- and population-level reproductive numbers, both of which are set to 2 in this example.
Parameters are chosen to reflect the West African Ebola outbreak \citep{who2014ebola}.}
\label{fig:indpop}
\end{figure}

Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence}; 
making this distinction clear will be particularly useful when we discuss spatial components later (see \fref{indpop}).
An individual-level intrinsic infection kernel $k(\tau; a)$ describes the rate at which an infected individual with ``aspect'' $a$ makes ``infectious contacts'' (contacts which will cause infection if the contactee is susceptible).
Individual aspects may represent variation in the course of infection (e.g., duration of latent and infectious periods) and the level of infectiousness, which can depend both on biological infectiousness as well as contact patterns.
Hereafter, we use $t, s$ to represent calendar time and $\tau$ to
represent time since infection.

@@

In a homogeneously mixing population, current disease incidence at time $t$, $i(t)$, is the product of the current infectiousness of individuals infected in the past and the current proportion of the population susceptible, $S(t)$.
\begin{equation}
i(t) = S(t) \int K(s) i(t-s)\,\mathrm{d}s = \RR_0 S(t) \int g(s) i(t-s)\,\mathrm{d}s.
\end{equation}
This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
Over a period of time where the susceptible proportion remains approximately constant ($S(t) \approx S(0)$), we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between the initial exponential growth rate $r$ and the initial reproductive number $\Rini = \Ro S(0)$:
\begin{equation}
\frac{1}{\Rini} = \int g(s) \exp(-r s)\,\mathrm{d}s.
\label{eq:euler}
\end{equation}
Under the homogeneous mixing assumption, the intrinsic generation-interval distribution $g(\tau)$ provides the correct link between $r$ and $\Rini$.

@@

Realized generation intervals can be measured either forward (from the perspective of a cohort of infectors) or backward (from the perspective of a cohort of infectees) in time \citep{champredon2015intrinsic, britton2019estimation}.
The forward generation-interval distribution $f_t(\tau)$ describes the infection time of \emph{infectees} caused by a cohort infectors who were infected at time $t$.
Similarly, the backward generation-interval distribution $b_t(\tau)$ describes the infection time of \emph{infectors} for a cohort of infectees who were infected at time $t$.
For a single infector-infectee pair, both backward and forward \emph{measurements} should give the identical generation interval.
Therefore, the total number of infections that occurs at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
\begin{equation}
\mathcal R_{\rm c}(t) i(t) f_t(\tau) = i(t+\tau) b_{t+\tau}(\tau).
\label{eq:equivalence}
\end{equation}
Here, $\mathcal R_{\rm c}(t)$ represents the case reproductive number, which is defined as the average number of secondary cases caused by a primary case infected at time $t$ over the course of their infection \citep{fraser2007estimating}.

In a homogeneously mixing population, the forward and backward generation-interval distributions can be calculated exactly.
The density of new infections occurring at time $t+\tau$ caused by infectors who were infected at time $t$ is given by
\begin{equation}
i_{t}(t+\tau) = \RR_0 i(t) g(\tau) S(t+\tau)
\end{equation}
As shown in \cite{champredon2015intrinsic}, the forward generation-interval distribution, $f_t(\tau)$, is proportional to $i_{t+\tau}(t)$:
\begin{equation}
f_t(\tau) = \frac{g(\tau) S(t+\tau)}{\int_0^\infty g(x) S(t+x) \,\mathrm{d}x}.
\label{eq:forward}
\end{equation}
In this case, the initial forward generation-interval distribution $f_0(\tau)$ during the exponential growth phase (when $S(t) \approx S(0)$) is equivalent to the intrinsic generation-interval distribution $g(\tau)$ and, therefore, provides the correct link between $r$ and \Rini.
Likewise, the density of new infections occurring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
\begin{equation}
i_{t-\tau}(t) = \RR_0 i(t-\tau) g(\tau) S(t)
\end{equation}
The backward generation-interval distribution, $b_t(\tau)$, is proportional to $i_{t-\tau}(t)$:
\begin{equation}
b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^\infty i(t-x) g(x)\,\mathrm{d}x}.
\end{equation}
Substituting $\mathcal R_{\rm c}(t) = \mathcal R_0 \int_0^\infty g(x) S(t+x) \,\mathrm{d}x$ confirms that \eref{equivalence} holds for this model.

During an ongoing epidemic, generation intervals cannot be measured for infection events that have not happened yet.
This effect is called ``right-truncation''.
Therefore, even if we aggregate all realized generation intervals by identifying who infected whom through contact tracing (assuming that infection events are observable), their mean will be shorter than the intrinsic generation intervals.
The aggregated generation-interval distribution, by definition, is a weighted average of backward generation-interval distributions (weighted by incidence) up until calendar time $t$:
\begin{equation}
a_t(\tau) \propto \int_{-\infty}^t i(s) b_s(\tau) \,\mathrm{d}s.
\end{equation}
The aggregated generation-interval distribution can be also equivalently expressed in terms of forward generation-interval distributions (weighted by incidence and case reproductive number):
\begin{equation}
a_t(\tau) \propto \int_{-\infty}^{t-\tau} \mathcal R_{\rm c}(s) i(s) f_s(\tau) \,\mathrm{d}s.
\end{equation}
\eref{equivalence} confirms that both expressions are identical.

For a single outbreak, the mean aggregated generation interval will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
There are two reasons for this phenomenon.
First, longer generation intervals are more likely than short intervals to be missed due to right truncation.
In particular, if we assume that the initial forward generation-interval distribution remains constant ($f_t(\tau) \approx f_0(\tau)$) when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
the initial aggregated (or backward) generation-interval distribution is just the initial forward generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
\begin{equation}
a_0(\tau) = b_0(\tau) \propto f_0(\tau) \exp(-r\tau).
\label{eq:exp}
\end{equation}
A deterministic simulation confirms that the aggregated generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
Second, the decreasing number of susceptibles over the course of an epidemic makes long infectious contacts less likely to result in infection \citep{champredon2015intrinsic}.
Overall, we therefore expect naively using the aggregated generation-interval distribution to underestimate the initial reproductive number.

@@

We first consider the infection process from an ``egocentric'' point of view, taking into account infectious contacts made by a single infector.
We define the egocentric kernel as the rate at which secondary infections are realized by a single primary case with aspect $a$ in the absence of other infectors:
\begin{equation}
\hat{k}(\tau; a) = k(\tau; a) \exp \left(- \delta(a) \int_0^\tau k(s; a) \,\mathrm{d}s\right),
\end{equation}
where $k(\tau; a)$ is the individual-level intrinsic kernel and $e^{- \delta(a) \int_0^\tau k(s; a) ds}$ is the probability that a susceptible acquaintance has not yet been contacted by the focal individual.
The dilution term, $\delta(a)$, models how contacts are distributed among the acquaintances.

Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \cite{trapman2016inferring}.
In this case, the intrinsic infectiousness of an individual $\RR(a) = \int k(s; a) \,\mathrm{d}s$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
for simplicity, we do not pursue these directions here.

The population-level egocentric kernel is found by integrating the individual-level kernel over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) \,\mathrm{d}a,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of the basic reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(s)\,\mathrm{d}s}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
Finally, the initial exponential growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\Rego} = \int \hat{g}(\tau) \exp(-r \tau)\,\mathrm{d}\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\Rego$ will be smaller than $\Rini$ estimated from the intrinsic distribution;
this generation-interval-based argument provides an alternative biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{diagram.pdf}
\caption{
\textbf{A summary of spatiotemporal effects on generation intervals.}
The intrinsic generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case.
The realized (forward) generation-interval distribution describes the time between actual infection events -- egocentric, local, and global depletion of the susceptible pool reduces the mean realized generation intervals.
The aggregated generation-interval distribution reflects the contact-tracing process during an ongoing epidemic and is subject to right truncation bias.
Our goal is to estimate the initial forward generation-interval distribution from the initial aggregated generation-interval distribution, measured during the exponential growth period.
}
\label{fig:diagram}
\end{figure}

@@

As expected, estimating the reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution (\eref{conditional}) address this problem only partially, as they do not account for indirect (local) spatial effects. 
Direct estimates based on the aggregated generation intervals from contact tracing (via Euler-Lotka) severely underestimate the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical estimates as well as the estimates based on the empirical initial forward generation-interval distribution on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimate of the growth rate.
For smaller values of \RR, we expect the differences to become smaller.
In the Appendix, we present the same figure using smaller \RR\ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
Overall, our qualitative conclusions do not change.

@@

Realized generation intervals that have been aggregated over the course of an epidemic are subject to right-truncation -- it is not possible to trace individuals that have not been infected yet.
The aggregated distributions can be thought of as averages of ``backward'' generation intervals (measured by looking at infectors of a cohort of individuals infected at the same time) \citep{champredon2015intrinsic, kenah2008generation, nishiura2010time, tomba2010some, britton2019estimation}.
During an ongoing outbreak, the aggregated generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-truncation.
Early in the outbreak, aggregated intervals are expected to match the early-outbreak backward intervals.
Near the end of an outbreak, the effect of right-truncation becomes negligible but the aggregated generation intervals are still shorter on average than intrinsic generation intervals, because of depletion of the susceptible population.

@@

We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
we previously showed that in many cases a gamma approximation (requiring estimation of only two parameters) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the estimated gamma distribution may ``look'' indistinguishable from the true shape of the intrinsic generation-interval distribution (derived from the SEIR model).
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Rini whereas narrower generation-interval distribution leads to higher \Rini) and, therefore, may have small effects on the overall estimate of $\Rini$ (Appendix A.4).

>>>>

At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
While this definition is widely used in the literature, it is not directly related to a population-level \emph{distribution}.
There are important distinctions to be made when defining generation-interval distributions at the population-level. The \emph{intrinsic} generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case \citep{champredon2015intrinsic}.
On the other hand, \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
Since some infectious contacts will be made with non-susceptible people, and thus not result in infection, realized distributions can differ systematically from the intrinsic distribution.

The shape of the realized generation-interval distribution depends on the reference time and perspective \citep{champredon2015intrinsic,kenah2008generation, nishiura2010time, tomba2010some, britton2019estimation}.
When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual. 
Thus, ``backward'' generation intervals, which look at a cohort of infectees and ask when their infectors were infected, will be shorter on average than intrinsic generation intervals -- the converse is true when an epidemic is subsiding \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
Likewise, we can define ``forward'' generation intervals, which look at a cohort of infectors and ask when their infectees were infected.
Mean forward generation intervals tends to decrease over the course of an epidemic due to susceptible depletion \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic}.

Realized generation intervals are also affected by spatial structure.
In a population that does not mix homogeneously, susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population.
This means that infectious contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already).
As a result, realized generation intervals (from the perspective of an infector) will typically have a shorter mean than the intrinsic generation-interval distribution in a non-homogeneous population.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller than would be predicted from homogeneous mixing.

In practice, realized generation intervals are often difficult to measure for many diseases, because it is difficult to observe when individuals become infected;
in most cases, observable events are clinical (e.g., onset of symptoms).
There are some exceptions: for example, generation intervals are commonly measured directly through contact tracing for canine rabies, where infection events are bites \citep{hampson2009transmission}.
Intervals between observed disease progression events (commonly, onset of signs or symptoms) are called serial intervals \citep{svensson2007note}.
Serial intervals are in many ways similar to generation intervals, but there are also complexities in their use  \citep{britton2019estimation}.
We will not address these complexities here.

While an epidemic is ongoing, realized generation intervals, at least in theory, can be measured by identifying who infected whom and when, and aggregated to form a single distribution.
We typically want to try to make inference based on this aggregated distribution --- that is, on all available data that have been gathered since the beginning of an epidemic.
These aggregated measurements are ``truncated'' because we do not know what happens after the time of last observation.
The distribution of these truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded in time to be observed.
We therefore predict that removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which contains information about the population structure and allows us to correctly infer the initial reproductive number from the initial exponential growth rate.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{individual_and_population.pdf}
\caption{\textbf{Comparison of individual- and population-level kernels.}
(Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days;
this represents an individual realization of a random process.
(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 days and 5 days, respectively;
this represents a population average of a random process.
Shaded areas under the curves are equal to individual- and population-level reproductive numbers, both of which are set to 2 in this example.
Parameters are chosen to reflect the West African Ebola outbreak \citep{who2014ebola}.}
\label{fig:indpop}
\end{figure}

Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence}; 
making this distinction clear will be particularly useful when we discuss spatial components later (see \fref{indpop}).
An individual-level intrinsic infection kernel $k(\tau; a)$ describes the rate at which an infected individual with ``aspect'' $a$ makes ``infectious contacts'' (contacts which will cause infection if the contactee is susceptible).
Individual aspects may represent variation in the course of infection (e.g., duration of latent and infectious periods) and the level of infectiousness, which can depend both on biological infectiousness as well as contact patterns.
Hereafter, we use $t, s$ to represent calendar time and $\tau, x$ to
represent time since infection.

@@

In a homogeneously mixing population, current disease incidence at time $t$, $i(t)$, is the product of the current infectiousness of individuals infected in the past and the current proportion of the population susceptible, $S(t)$.
\begin{equation}
i(t) = S(t) \int K(\tau) i(t-\tau)\,\mathrm{d}\tau = \RR_0 S(t) \int g(\tau) i(t-\tau)\,\mathrm{d}\tau.
\end{equation}
This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{roberts2007model, wallinga2007generation, heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, Champredon2018equivalence}.
Over a period of time where the susceptible proportion remains approximately constant ($S(t) \approx S(0)$), we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between the initial exponential growth rate $r$ and the initial reproductive number $\Rini = \Ro S(0)$:
\begin{equation}
\frac{1}{\Rini} = \int g(\tau) \exp(-r \tau)\,\mathrm{d}\tau.
\label{eq:euler}
\end{equation}
Under the homogeneous mixing assumption, the intrinsic generation-interval distribution $g(\tau)$ provides the correct link between $r$ and $\Rini$.

@@

Realized generation intervals can be measured either forward (from the perspective of a cohort of infectors) or backward (from the perspective of a cohort of infectees) in time \citep{champredon2015intrinsic, britton2019estimation}.
The forward generation-interval distribution $f_t(\tau)$ describes the infection time of \emph{infectees} caused by a cohort infectors who were infected at time $t$.
Similarly, the backward generation-interval distribution $b_t(\tau)$ describes the infection time of \emph{infectors} for a cohort of infectees who were infected at time $t$.
For a single infector-infectee pair, both backward and forward \emph{measurements} should give the identical generation interval.
Therefore, the density of new infections occurring at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
\begin{equation}
\mathcal R_{\rm c}(t) i(t) f_t(\tau) = i(t+\tau) b_{t+\tau}(\tau).
\label{eq:equivalence}
\end{equation}
Here, $\mathcal R_{\rm c}(t)$ represents the case reproductive number, which is defined as the average number of secondary cases caused by a primary case infected at time $t$ over the course of their infection \citep{fraser2007estimating}.

In a homogeneously mixing population, the forward and backward generation-interval distributions can be calculated exactly.
The density of new infections occurring at time $t+\tau$ caused by infectors who were infected at time $t$ is given by
\begin{equation}
i_{t}(t+\tau) = \RR_0 i(t) g(\tau) S(t+\tau)
\end{equation}
As shown in \cite{champredon2015intrinsic}, the forward generation-interval distribution, $f_t(\tau)$, is proportional to $i_{t}(t+\tau)$:
\begin{equation}
f_t(\tau) = \frac{g(\tau) S(t+\tau)}{\int_0^\infty g(x) S(t+x) \,\mathrm{d}x}.
\label{eq:forward}
\end{equation}
In this case, the initial forward generation-interval distribution $f_0(\tau)$ during the exponential growth phase (when $S(t) \approx S(0)$) is equivalent to the intrinsic generation-interval distribution $g(\tau)$ and, therefore, provides the correct link between $r$ and \Rini.
Likewise, the density of new infections occurring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
\begin{equation}
i_{t-\tau}(t) = \RR_0 i(t-\tau) g(\tau) S(t)
\end{equation}
The backward generation-interval distribution, $b_t(\tau)$, is proportional to $i_{t-\tau}(t)$:
\begin{equation}
b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^\infty i(t-x) g(x)\,\mathrm{d}x}.
\end{equation}
Substituting $\mathcal R_{\rm c}(t) = \mathcal R_0 \int_0^\infty g(\tau) S(t+\tau) \,\mathrm{d}\tau$ confirms that \eref{equivalence} holds for this model.

During an ongoing epidemic, generation intervals cannot be measured for infection events that have not happened yet.
This effect is called ``right-truncation''.
Therefore, even if we aggregate all realized generation intervals by identifying who infected whom through contact tracing (assuming that infection events are observable), their mean will be shorter than the mean intrinsic generation interval.
The aggregated generation-interval distribution, by definition, is a weighted average of backward generation-interval distributions (weighted by incidence) up until calendar time $t$:
\begin{equation}
a_t(\tau) \propto \int_{-\infty}^t i(s) b_s(\tau) \,\mathrm{d}s.
\end{equation}
The aggregated generation-interval distribution can be also expressed equivalently in terms of forward generation-interval distributions (weighted by incidence and case reproductive number):
\begin{equation}
a_t(\tau) \propto \int_{-\infty}^{t-\tau} \mathcal R_{\rm c}(s) i(s) f_s(\tau) \,\mathrm{d}s.
\end{equation}
\eref{equivalence} confirms that both expressions are identical.

For a single outbreak, the mean aggregated generation interval will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
There are two reasons for this phenomenon.
First, longer generation intervals are more likely than short intervals to be missed due to right truncation.
In particular, if we assume that the initial forward generation-interval distribution remains constant ($f_t(\tau) \approx f_0(\tau)$) when an epidemic is growing exponentially ($i(t) \approx i(0) \exp(rt)$ and $\mathcal R_{\rm c}(t) \approx \mathcal R_{\rm c}(0)$),
the initial aggregated (or backward) generation-interval distribution is just the initial forward generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
\begin{equation}
a_0(\tau) = b_0(\tau) \propto f_0(\tau) \exp(-r\tau).
\label{eq:exp}
\end{equation}
A deterministic simulation confirms that the aggregated generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
Second, the decreasing number of susceptibles over the course of an epidemic makes long infectious contacts less likely to result in infection \citep{champredon2015intrinsic}.
Overall, we therefore expect naively using the aggregated generation-interval distribution to underestimate the initial reproductive number.

@@

We first consider the infection process from an ``egocentric'' point of view, taking into account infectious contacts made by a single infector.
We define the egocentric kernel as the rate at which secondary infections are realized by a single primary case with aspect $a$ in the absence of other infectors:
\begin{equation}
\hat{k}(\tau; a) = k(\tau; a) \exp \left(- \delta(a) \int_0^\tau k(x; a) \,\mathrm{d}x\right),
\end{equation}
where $k(\tau; a)$ is the individual-level intrinsic kernel and $e^{- \delta(a) \int_0^\tau k(x; a)  \,\mathrm{d}x}$ is the probability that a susceptible acquaintance has not yet been contacted by the focal individual.
The dilution term, $\delta(a)$, models how contacts are distributed among the acquaintances.

Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \cite{trapman2016inferring}.
In this case, the intrinsic infectiousness of an individual $\RR(a) = \int k(\tau; a) \,\mathrm{d}\tau$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, the contact rate $\lambda$, and the duration of infectious period; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
for simplicity, we do not pursue these directions here.

The population-level egocentric kernel is found by integrating the individual-level kernel over individual variations:
\begin{equation}\label{eq:ego}
\hat{K}(\tau) = \int \hat{k}(\tau; a) f(a) \,\mathrm{d}a,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
Essentially, the population-level egocentric kernel accounts for the probability that a susceptible individual has not been infected by the focal individual.
Trapman \textit{et al.} \cite{trapman2016inferring} used this same kernel (also assuming a constant per-pair contact rate) to study the effect of network structure on the estimate of the basic reproductive number.
The population-level egocentric generation-interval distribution is:
\begin{equation}
\hat{g}(\tau) = \frac{\hat{K}(\tau)}{\int \hat{K}(x)\,\mathrm{d}x}.
\label{eq:conditional}
\end{equation}
The population-level egocentric generation-interval distribution describes the distribution of times at which secondary infections are realized from an \emph{average} infected-susceptible pair; for convenience, we will often omit ``population-level''.
Finally, the initial exponential growth rate and the egocentric reproductive number are linked by the egocentric generation-interval distribution (and the Euler-Lotka equation) \citep{trapman2016inferring}:
\begin{equation}
\frac{1}{\Rego} = \int \hat{g}(\tau) \exp(-r \tau)\,\mathrm{d}\tau.
\label{eq:egorR}
\end{equation}
As the egocentric distribution always has a shorter mean than the intrinsic distribution, $\Rego$ will be smaller than $\Rini$ estimated from the intrinsic distribution;
this generation-interval-based argument provides an alternative biological interpretation for the result presented by \cite{trapman2016inferring}.
% Similarly, \Rhat\ will be greater than the \emph{true} reproductive number \RR\, since it does not account for depletion of susceptibles by other routes.

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{diagram.pdf}
\caption{
\textbf{A summary of spatiotemporal effects on generation intervals.}
The intrinsic generation-interval distribution describes the expected time distribution of infectious contacts made by a primary case.
The realized (forward) generation-interval distribution describes the time between actual infection events --- egocentric, local, and global depletion of the susceptible pool reduces the mean realized generation intervals.
The aggregated generation-interval distribution reflects the contact-tracing process during an ongoing epidemic and is subject to right truncation bias.
Our goal is to estimate the initial forward generation-interval distribution from the initial aggregated generation-interval distribution, measured during the exponential growth period.
}
\label{fig:diagram}
\end{figure}

@@

As expected, estimating the reproductive number based on the intrinsic generation-interval distribution overestimates the empirical reproductive number;
estimates based on the egocentric generation-interval distribution (\eref{conditional}) address this problem only partially, as they do not account for indirect (local) spatial effects. 
Direct estimates based on the aggregated generation intervals from contact tracing (via Euler-Lotka) severely underestimate the empirical estimates.
While both population- and individual-level corrections provide similar estimates to the empirical estimates as well as the estimates based on the empirical initial forward generation-interval distribution on average,
population-level estimates are more variable as they are more sensitive to outliers in generation intervals and our estimates of the initial exponential growth rate.
For smaller values of $\Rini$, we expect the differences to become smaller.
In the Appendix, we present the same figure using smaller $\Rini$ (Appendix A.1) and using Erlang-distributed latent periods (Appendix A.2), which better corresponds to Ebola.
Overall, our qualitative conclusions do not change.

@@

Realized generation intervals that have been aggregated over the course of an epidemic are subject to right-truncation -- it is not possible to trace individuals that have not been infected yet.
The aggregated distributions can be thought of as averages of ``backward'' generation intervals (measured by looking at infectors of a cohort of individuals infected at the same time) \citep{champredon2015intrinsic, kenah2008generation, nishiura2010time, tomba2010some, britton2019estimation}.
During an ongoing outbreak, the aggregated generation-interval distribution will always have a shorter mean than the intrinsic-interval distribution due to right-truncation.
Early in the outbreak, the initial aggregated intervals are expected to match the initial backward intervals.
Near the end of an outbreak, the effect of right-truncation becomes negligible but the aggregated generation intervals are still shorter on average than intrinsic generation intervals, because of depletion of the susceptible population.

@@

We compare two methods for estimating the initial forward generation-interval distribution and assume that the initial forward generation-interval distribution follows a gamma distribution.
The gamma approximation of the generation interval distribution has been widely used due to its simplicity \citep{mcbryde2009early, nishiura2009transmission, roberts2011early, trichereau2012estimation, nishiura2015theoretical};
we previously showed that a gamma approximation (requiring estimation of only two parameters) can be sufficient to understand the role of generation-interval distributions in linking $r$ and $\Rini$ for Ebola, rabies, and measles \citep{park2019practical}.
However, further investigation of our methods suggests that making a wrong distributional assumption can lead to biased estimates of the mean and CV of a generation-interval distribution (Appendix A.4), even though the estimated gamma distribution may ``look'' indistinguishable from the true shape of the intrinsic generation-interval distribution (derived from the SEIR model).
These results are particularly alarming because it is impossible to know the true shape of the generation-interval distribution for real diseases.
Nonetheless, biases in the parameter estimates of a generation-interval distribution may have opposite effects on the estimate of \Rini (e.g., shorter mean generation interval leads to lower \Rini whereas narrower generation-interval distribution leads to higher \Rini) and, therefore, may have small effects on the overall estimate of $\Rini$ (Appendix A.4).
